\pkg{evtree}: Evolutionary Learning of Globally Optimal Classification and Regression Trees in \proglang{R}

### 2011-11-07 pkg/evtree/inst/doc/evtree.Rnw
fixed typo
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@253 255f4249-2263-4417-b1ea-8498779e27a7
fixed a few typos
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@269 255f4249-2263-4417-b1ea-8498779e27a7

<<setup, echo=FALSE>>=
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("rpart")
library("evtree")
data("BBBClub", package = "evtree")
cache <- TRUE
@

@@

To show why forward-search recursive partitioning algorithms typically
lead to globablly suboptimal solutions, their parameter spaces and optimization
problems are presented and contrasted in a unified notation. Although all
arguments hold more generally, only binary tree models with some maximum number
of splits $M_{\max}$ are considered. Both restrictions make the notation somewhat
simpler while not really restricting the problem: (a)~Multiway splits are equivalent
to a sequence of binary splits in predictions and number of resulting subsamples.
(b)~The maximal size of the tree is always limited by the number of observations
in the learning sample.

@@

The search for a valid split rule is conducted as in see Section~\ref{init}
for a maximum of $P$~iterations. 
In cases where no valid split rule can be assigned to the internal node at
position~$r$, the search for a valid split rule is carried out on another
randomly selected terminal node. If, after 10~attempts, no valid split rule
can be found, then ${\theta}_{i+1} = {\theta}_{i}$. Otherwise, the
set of parameters in iteration $i+1$ are given by ${\theta}_{i+1}=
{\theta}_i \cup (v_{r},s_{r})$.

@@

\textit{Minor split rule mutation} is similar to the \textit{major split rule
mutation} operator, but it does not alter $v_r$ and only changes the split point
$s_r$ by a minor degree. If $X_{v_r}$ is numerical or ordinal the split point
$s_r$ is changed by a non-zero number of unique values of $X_{v_r}$. In cases
where $X_{v_r}$ has less then $20$ unique values, the split point is change to
to the next larger, or the next lower, unique value of $X_{v_r}$. Otherwise,
$s_r$ is randomly shifted by a number of unique values that is not larger than
$10\%$ of the range of unique values of $X_{v_r}$. If $X_{v_r}$ is a nominal
variable, with less than $20$ categories, one of the categories is randomly
modified. Otherwise, at least one and at most $10\%$ of the variable's
categories are changed. In cases where subsequent nodes become invalid, further
split points are searched that preserve the tree's topology. After five
non-successful attempts at finding a topology preserving split point, the
non-valid nodes are pruned.

@@

\textit{Crossover} exchanges, randomly selected, subtrees between two trees. Let
${\theta}_{i}^1$ and ${\theta}_{i}^2$ be the two trees
chosen from the population for crossover. First, an internal nodes $r_1$ and $r_2$
are selected randomly from ${\theta}_{i}^1$ and ${\theta}_{i}^2$, respectively.
Let $\mathrm{sub}1({\theta}_i^j, r_j)$ denote the subtree of $\theta_j$ rooted
by $r_j$ ($j=1,2$), i.e., the tree containing $r_j$ and its descendant nodes.
Then, the complementary part of ${\theta}_{i}^j$ can be defined as 
$\mathrm{sub}2({\theta}_{i}^j, r_j) = {\theta}_{i}^j \setminus \mathrm{sub}1({\theta}_{i}^j, r_j)$. 
The crossover operator creates two new trees
${\theta}_{i+1}^1 = \mathrm{sub}2({\theta}_{i}^1, r_1) \cup \mathrm{sub}1({\theta}_{i}^2, r_2)$ and
${\theta}_{i+1}^2 = \mathrm{sub}2({\theta}_{i}^2, r_2) \cup \mathrm{sub}1({\theta}_{i}^1, r_1)$.
If the crossover creates some invalid nods in either one of the new trees ${\theta}_{i+1}^1$
or ${\theta}_{i+1}^2$, these are omitted.

@@

Figure~\ref{fig:benchmark} (lower panels) summarizes the relative performance of \code{evtree} and \code{ctree}. For 15 out of 17 datasets \code{evtree} shows a better predictive performance. The algorithms' performances  is significantly worse on the \textit{MEL0101} dataset, where the average misclassifiation rate of \code{ctree} is $5.6\%$ lower. However, on this dataset, \code{ctree} models are on average $86.5\%$ larger than \code{evtree} models. The relative complexity of \code{evtree} models is significantly smaller for 15 and larger for 1 dataset.

>>>>

<<setup, echo=FALSE>>=
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("rpart")
library("evtree")
data("BBBClub", package = "evtree")
cache <- FALSE
@

@@

To show why forward-search recursive partitioning algorithms typically
lead to globablly suboptimal solutions, their parameter spaces and optimization
problems are presented and contrasted in a unified notation. Although all
arguments hold more generally, only binary tree models with some maximum number
of terminal nodes $M_{\max}$ are considered. Both restrictions make the notation somewhat
simpler while not really restricting the problem: (a)~Multiway splits are equivalent
to a sequence of binary splits in predictions and number of resulting subsamples.
(b)~The maximal size of the tree is always limited by the number of observations
in the learning sample.

@@

The search for a valid split rule is conducted as in Section~\ref{init}
for a maximum of $P$~iterations. 
In cases where no valid split rule can be assigned to the internal node at
position~$r$, the search for a valid split rule is carried out on another
randomly selected terminal node. If, after 10~attempts, no valid split rule
can be found, then ${\theta}_{i+1} = {\theta}_{i}$. Otherwise, the
set of parameters in iteration $i+1$ are given by ${\theta}_{i+1}=
{\theta}_i \cup (v_{r},s_{r})$.

@@

\textit{Minor split rule mutation} is similar to the \textit{major split rule
mutation} operator, but it does not alter $v_r$ and only changes the split point
$s_r$ by a minor degree. If $X_{v_r}$ is numerical or ordinal the split point
$s_r$ is changed by a non-zero number of unique values of $X_{v_r}$. In cases
where $X_{v_r}$ has less then $20$ unique values, the split point is changed to
the next larger, or the next lower, unique value of $X_{v_r}$. Otherwise,
$s_r$ is randomly shifted by a number of unique values that is not larger than
$10\%$ of the range of unique values of $X_{v_r}$. If $X_{v_r}$ is a nominal
variable, with less than $20$ categories, one of the categories is randomly
modified. Otherwise, at least one and at most $10\%$ of the variable's
categories are changed. In cases where subsequent nodes become invalid, further
split points are searched that preserve the tree's topology. After five
non-successful attempts at finding a topology preserving split point, the
non-valid nodes are pruned.

@@

\textit{Crossover} exchanges, randomly selected, subtrees between two trees. Let
${\theta}_{i}^1$ and ${\theta}_{i}^2$ be the two trees
chosen from the population for crossover. First, two internal nodes $r_1$ and $r_2$
are selected randomly from ${\theta}_{i}^1$ and ${\theta}_{i}^2$, respectively.
Let $\mathrm{sub}1({\theta}_i^j, r_j)$ denote the subtree of $\theta_j$ rooted
by $r_j$ ($j=1,2$), i.e., the tree containing $r_j$ and its descendant nodes.
Then, the complementary part of ${\theta}_{i}^j$ can be defined as 
$\mathrm{sub}2({\theta}_{i}^j, r_j) = {\theta}_{i}^j \setminus \mathrm{sub}1({\theta}_{i}^j, r_j)$. 
The crossover operator creates two new trees
${\theta}_{i+1}^1 = \mathrm{sub}2({\theta}_{i}^1, r_1) \cup \mathrm{sub}1({\theta}_{i}^2, r_2)$ and
${\theta}_{i+1}^2 = \mathrm{sub}2({\theta}_{i}^2, r_2) \cup \mathrm{sub}1({\theta}_{i}^1, r_1)$.
If the crossover creates some invalid nods in either one of the new trees ${\theta}_{i+1}^1$
or ${\theta}_{i+1}^2$, these are omitted.

@@

Figure~\ref{fig:benchmark} (lower panels) summarizes the relative performance of \code{evtree} and \code{ctree}. For 15 out of 17 datasets \code{evtree} shows a better predictive performance. The algorithms' performances  is significantly worse on the \textit{MEL0101} dataset, where the mean squarred error of \code{ctree} is $5.6\%$ lower. However, on this dataset, \code{ctree} models are on average $86.5\%$ larger than \code{evtree} models. The relative complexity of \code{evtree} models is significantly smaller for 15 and larger for 1 dataset.
